\mainmatter

# Introducción (aka. About)

This is a *sample* book written in **Markdown**\index{markdown}. You can use anything that Pandoc's Markdown supports; for example, a math equation\index{ecuación} $a^2 + b^2 = c^2$ según @eddelbuettel2017ExtendingBrief.

Este primer capítulo numerado (00-realcap1.Rmd) después de sacarlo del index.Rmd. Se agregan los archivos complementarios del frontmatter que se numerarían en romanos y en el índice ver si aparecen...

## Usage

Each **bookdown**\index{bookdown} chapter\index{capítulo} is an .Rmd file, and each .Rmd file can contain one (and only one) chapter. A chapter *must* start with a first-level heading: `# A good chapter`, and can contain one (and only one) first-level heading.

Use second-level and higher headings within chapters like: `## A short section` or `### An even shorter section`.

The `index.Rmd` file is required\index{obligatorio}, and is also your first book chapter. It will be the homepage when you render the book \index{libro}.

## Render book

You can render the HTML version of this example book without changing anything:

1.  Find the **Build** pane in the RStudio IDE, and

2.  Click on **Build Book**, then select your output format, or select "All formats" if you'd like to use multiple formats from the same book source files.

Or build the book from the R console:

```{r, eval=FALSE}
bookdown::render_book()
```

To render this example to PDF \index{PDF} as a `bookdown::pdf_book`, you'll need to install XeLaTeX. You are recommended to install TinyTeX \index{TinyTeX}(which includes XeLaTeX): <https://yihui.org/tinytex/>.

## Preview book \index{vista previa}

As you work, you may start a local server to live preview this HTML book. This preview will update as you edit the book when you save individual .Rmd files. You can start the server in a work session by using the RStudio add-in "Preview book", or from the R console:

```{r eval=FALSE}
bookdown::serve_book()
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

Esta es la prueba en modo visual para insertar \index{insertar}directo una referencia solo con la arroba @wickham2019Advanced y los demás [@wickham2019Advanced; @xie2015]. En realidad no fununcia; sin embargo, en el modo visual, el atajo Ctrl+SHift+F8 abre el manejador de citas y permite seleccionar la(s) deseada(s) para insertar la bibkey [@R-base].

La primera prueba es con la base generada directo por Zotero \index{Zotero}(Export selected as BibTeX, @saldierna-calapiz2018Efectored).

@ahmad2016UsingFisher establece que @anscombe1973GraphsStatistical está totalmente equivocado al insertar directo desde Zotero y agregarla al archivo de bibliografía [@bravo2016Teachinghigher].

Lo que sigue se supone que es ***Markdown extended***:

Por ejemplo, la definición de términos

hábrase visto:

:   expresión choyera de asombro!

relación:

:   solo para acompletar el punto; ponemos mucho texto para ver como maneja el párrafo. As you work, you may start a local server to live preview this HTML book. This preview will update as you edit the book when you save individual .Rmd files. You can start the server in a work session by using the RStudio add-in "Preview book", or from the R console

También hay una lista de pendientes, parecido a la lista ordenada o no

-   [ ] Write the press release
-   [ ] Update the website
-   [ ] Contact the media

Así como el resalte de ==lo más importante en el texto== y ver si fununcia. Este marcador extendido para resaltar no es interpretado en **PDF** (por lo menos).

La referencia la ponemos a pie de página forzada [^index-1]. El formato para insertarlo es \^[texto del pie de página]; RMarkdown lo interpreta y lo deja numerado y estandarizado -como aparece en el *Rmd*-.

[^index-1]: <https://www.markdownguide.org/cheat-sheet>
